---
title: Health and Economic Impact of Severe Weather Events Across the USA - Exploring
  the NOAA Storm Database
output:
  html_document:
    df_print: paged
---

## Synopsis

This project is intented to target the exploration of the NOAA Storm Database to understand the impact on public health and overall economics of severe weather events, like tornados, drought, snow, rainfalls, in several categories. Two basic questions are adressed by this report:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

This repost will try to answer these questions showing relevant data and their graphical interpretion. Working software code will be available as well, to help with its reproducibility.

## Data Processing

```{r echo=TRUE, message=FALSE, comment=FALSE, results=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(reshape2)
```

The data for this project is available from this link

> https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2

```{r echo=TRUE, message=FALSE, cache=TRUE, comment=FALSE, results=FALSE}
dest_file <- "StormData.csv.bz2"
if (!file.exists(dest_file)) {
        url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(url, destfile = dest_file)
        download_timestamp <- timestamp()
        write(download_timestamp, file = "download_timestamp.txt")
}
download_timestamp <- read.delim("download_timestamp.txt", header = FALSE)
```

File *`r dest_file`* was downloaded on

> `r gsub("  ", " ", gsub("#| ?-+ ?", "", unlist(download_timestamp)))`

```{r cache=TRUE}
dataset <- read.csv(dest_file)
```

A sample from the original dataset is shown below
```{r echo=TRUE}
head(dataset)
```

The dataset has `r format(dim(dataset)[1], big.mark = ",")` records. 


The first step is to normalize the event type fields with the 48 different types recognized by the NOAA

```{r}
event_types <- list(
        c("thunderstorm wind", "thunderstorm|tstm"),
        c("storm surge/tide", "(surge|high.+tide)"),
        c("astronomical low tide", "astronomical|low +tide"),
        c("avalanche", "avalanc"),
        c("blizzard", "blizzard"),
        c("coastal flood", "coastal"),
        c("extreme cold/wind chill", "extreme.+chill"),
        c("cold/wind chill", "chill"),
        c("debris flow", "debris"),
        c("freezing fog", "free.+fog"),
        c("dense fog", "fog"),
        c("dense smoke", "smoke"),
        c("drought", "drought"),
        c("dust devil", "devil"),
        c("dust storm", "dust +storm"),
        c("excessive heat", "exc.+heat"),
        c("lakeshore flood ", "lakeshore"),
        c("flash flood", "flash"),
        c("flood", "flood"),
        c("frost/freeze", "frost|freeze"),
        c("funnel cloud", "funnel"),
        c("marine hail", "marine.+hail"),
        c("hail", "hail"),
        c("heat", "heat"),
        c("heavy rain", "rain"),
        c("heavy snow", "snow"),
        c("high surf", "surf"),
        c("high wind", "high.+wind"),
        c("hurricane (typhoon)", "hurricane|typhoon"),
        c("ice storm", "ice.+storm"),
        c("lake-effect snow", "lake"),
        c("lightning", "lightning"),
        c("marine high wind", "marine.+high"),
        c("marine strong wind", "marine.+strong"),
        c("marine thunderstorm wind", "marine +thunderstorm +wind"),
        c("rip current", "rip|current"),
        c("seiche", "seiche"),
        c("sleet", "sleet"),
        c("strong wind", "strong +wind"),
        c("tornado", "tornado"),
        c("tropical depression", "tropical.+depression"),
        c("tropical storm", "tropical.+storm"),
        c("tsunami", "tsunami"),
        c("volcanic ash", "volcanic|ash"),
        c("waterspout", "waterspout"),
        c("wildfire", "wildfire"),
        c("winter storm", "winter.+storm"),
        c("winter weather", "winter.+weather")
)
```

The second field of each element of the list above is a regular expression that will be used to identify the existing values in the dataset. Then, they will be changed to the official naming from the NOAA database

```{r cache=TRUE, comment=FALSE, message=FALSE, results=FALSE}
normalize_event_type <- function(data_set, event) {
        data_set[grep(event[2], data_set$EVTYPE, ignore.case = TRUE), 'EVTYPE'] <- event[1]
        data_set
}

dataset$EVTYPE <- as.character(unlist(dataset$EVTYPE))

for (event in event_types) {
        dataset <- normalize_event_type(dataset, event)
}
```

### Types of Events Most Harmful to Population Health

The relevant fields to answer this first question are
 
> EVTYPE &rarr; EventType (renamed)  
> FATALITIES &rarr; Fatalities (renamed)  
> INJURIES &rarr; Injuries (renamed)    

The initial data exploration will remove the records with non-relevant data (fatalities and injuries equal to zero) and will change the magnetude of both featues by dividing by 10<sup>3</sup>, so the presented data can be more readable.

```{r echo=TRUE, comment=FALSE, message=FALSE, results=FALSE}
health_impact_set <- dataset %>%
        select(EVTYPE, FATALITIES, INJURIES) %>%
        mutate(Fatalities = FATALITIES / 1e3,
               Injuries = INJURIES / 1e3,
               EventType = tolower(EVTYPE)) %>%
        filter(Fatalities != 0 | Injuries != 0) %>%
        select(EventType, Fatalities, Injuries)
```

Summing up the different events and extracting the top ten most impacting ones is the next step in the process

```{r comment=FALSE, message=FALSE, results=FALSE}
health_impacting_events <- aggregate(cbind(Fatalities, Injuries) ~ EventType, data = health_impact_set, sum)
top_ten_health_impacting_events <- health_impacting_events %>%
        mutate(Total = Fatalities + Injuries) %>%
        top_n(10)
most_health_impacting_events <- top_ten_health_impacting_events %>%
        select(EventType, Fatalities, Injuries) %>%
        melt(id = "EventType", variable.name = "Casualties")
```

### Types of Events Most Harmful to the Economy

The relevant fields to answer the second question are
  
> EVTYPE &rarr; EventType (renamed)  
> PROPDMG &rarr; PropDmg (renamed, property damage)  
> PROPDMGEXP &rarr; (property damage exponent)    
> CROPDMG &rarr; CropDmg (renamed, crop damage)  
> CROPDMGEXP &rarr; (crop damage exponent)  

The exponent values (used to multiply the values in PROPDMG and CROPDMG) in the fields PROPDMGEXP and CROPDMGEXP will be replaced following the logic

> k or K &rarr; 1x10<sup>3</sup>  
> m or M &rarr; 1x10<sup>6</sup>  
> b or B &rarr; 1x10<sup>9</sup>  
> h or H &rarr; 1x10<sup>2</sup>  
> 0, +, ?, - or empty &rarr; 1  
> any number will be as is

as seen in the function *multiplier*

```{r echo=TRUE}
multiplier <- function(type) {
        mul <- tolower(as.character(type))
        switch(ifelse(mul == '', '1', mul),
               'b' = 1e9,
               'm' = 1e6,
               'k' = 1e3,
               'h' = 1e2,
               '+' = 1,
               '-' = 1,
               '0' = 1,
               '?' = 1,
               as.numeric(type))
}
```

Once again, the initial preparation of the data removes double zero values (from PRODMG and CROPDMG)

```{r comment=FALSE, message=FALSE, results=FALSE}
economic_impact_set <- dataset %>%
        select(EVTYPE, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP) %>%
        mutate(EventType = EVTYPE,
               PropDmg = PROPDMG,
               CropDmg = CROPDMG) %>%
        filter(PropDmg != 0 | CropDmg != 0) %>%
        select(EventType, PropDmg, CropDmg, PROPDMGEXP, CROPDMGEXP)
```

The exponent fields are changed by the *multiplier* function and the final result is divided by 10<sup>9</sup>, for better reading in the display table

```{r comment=FALSE, message=FALSE, results=FALSE}
economic_impact_set$PropDmgMul <- economic_impact_set$PROPDMGEXP %>%
        sapply(multiplier)
economic_impact_set$CropDmgMul <- economic_impact_set$CROPDMGEXP %>%
        sapply(multiplier)
economic_impact_set <- economic_impact_set %>%
        mutate(Property = PropDmg * PropDmgMul / 1e9,
               Crop = CropDmg * CropDmgMul / 1e9) %>%
        select(EventType, Property, Crop)
```

The relevant fields, `Property` and `Crop`, are summed up, so the top ten most impacting event can be highlighted

```{r comment=FALSE, message=FALSE, results=FALSE}
econ_agg <- aggregate(cbind(Property, Crop) ~ EventType, data = economic_impact_set, sum)
top_ten_econ_events <- econ_agg %>%
        mutate(Total = Property + Crop) %>%
        top_n(10)
most_economic_impacting_events <- top_ten_econ_events %>%
        select(EventType, Property, Crop) %>%
        melt(id = "EventType", variable.name = "Damage")
```

## Results

### Human Health Impact

The top ten event types that impact human health are shown below. By far, tornadoes are the most impacting severe weather event, with `r format(max(top_ten_health_impacting_events$Total)/sum(health_impacting_events[, c('Fatalities', 'Injuries')]) * 100, digits = 3, nsmall = 1)`% of the total number of casualties. Values are in thousands
```{r}
top_ten_health_impacting_events %>%
        arrange(desc(Total))
```

```{r}
g <- most_health_impacting_events %>% ggplot(aes(fill = Casualties, y = value, x = reorder(EventType, value)))
caption <- "Fatalities and injuries caused by severe weather events. Values are in thousands"
g + geom_bar(stat = "identity") +
        ggtitle("Number of Casualties per Event Type") +
        ylab(paste("Casualties (in thousands)", caption, sep = "\n\n")) +
        xlab("Event Type") +
        coord_flip()
```

### Economic Impact

The highest economic impact is caused by floods, while the other 9 top severe weather events are trailing behind with, at most, roughly half of the floods' cost. Values are in US$ Billions

```{r}
top_ten_econ_events %>%
        arrange(desc(Total))
```

```{r}
g <- most_economic_impacting_events %>% ggplot(aes(fill = Damage, x = reorder(EventType, value), value))
caption <- "Economic impact of severe weather events. Values are in US$ Billions"
g + geom_bar(stat = "identity") +
        ggtitle("Economic Cost per Event Type") +
        ylab(paste("Cost (in US$ billions)", caption, sep = "\n\n")) +
        xlab("Event Type") +
        coord_flip()
```
